---
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
    code_download: true
title: "Can we predict one's biological age?"
subtitle: "Introduction to Linear Modelling"
date: "September 16th, 2022"
author:
  - name: Joao Malato
  - name: Nuno Sepulveda
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = TRUE,
                      message = TRUE, 
                      fig.align = "center",
                      fig.asp = 0.618,
                      fig.width = 10,
                      dpi = 200, 
                      out.width = "75%")
```


# Libraries {-}

```{r libraries, message=FALSE, warning=FALSE}
library(here)       # easy directory management, as here("YOUR/FILE/PATH")
library(data.table) # fast data wrangling
library(magrittr)   # use of pipe operator `%>%`
library(ggplot2)    # graphics
library(nortest)    # normality assessment test
```


# Load data

We will be analysing a subset of the whole blood data from 656 human individuals, made available by [Hannun et al. (2013)](https://www.sciencedirect.com/science/article/pii/S1097276512008933?via%3Dihub).

```{r}
# base R
mdna <- read.csv(here("data/methDNA_data_Hannun_etal_2013.csv"), header = TRUE)

# alternative: using data.table's fread()
# mdna <- fread(here("data/methDNA_data_Hannun_etal_2013.csv"), header = TRUE)
```

```{r}
# look at first 8 cols from the data
head(mdna[, 1:8])
```

Variables:

* `GEO.ID`: Unique identifier for each one of the `r nrow(mdna)` participants
* `Age`: Age of participants, in years
* `Gender`: Gender of each individual, binary values for female (`F`) and male (`M`)
* `Ethnicity`: Ethnic group of each individual, binary identifying European participants (`Caucasian`) and Mexican participants (`Hispanic`)
* `cgXXXXXXXX`: CpG marker, each one indicating the methylation values, in percentage

```{r}
# get a vector with all column names with their pattern starting with "cg" (i.e., the names of all CpG probes)
vec_probe <- grep("^cg", colnames(mdna), value = TRUE)
```


# Quick EDA

Things to look for:

1. Look at the population-related variables
    + Are there sampling differences between gender and/or ethnicity?
2. Look at age distribution
    + Does age change with gender/ethnicity?
3. Look at methylation levels
    + Should we transform the data?
    + Are the values correlated with age?
    + Could these values by affected by gender/ethnicity?


## Demographics and age

```{r}
# Look through the group categories
table(mdna$Gender)
table(mdna$Ethnicity)

# gender and ethnicity
table(mdna$Ethnicity, mdna$Gender)

# proportions
round(prop.table(table(mdna$Ethnicity, mdna$Gender), margin = 2), 2)
round(prop.table(table(mdna$Ethnicity, mdna$Gender), margin = 1), 2)

# Pearson's Chi-squared test
chisq.test(table(mdna$Gender, mdna$Ethnicity))
```


```{r}
# Gender and Age
boxplot(mdna$Age ~ mdna$Gender)

quantile(subset(mdna, Gender == "F")$Age)
quantile(subset(mdna, Gender == "M")$Age)

# Mann-Whitney U test
wilcox.test(mdna$Age ~ mdna$Gender)
```

```{r}
# ggplot2 alternatives

# boxplot:
# ggplot(mdna, aes(x = Gender, y = Age)) +
#   geom_boxplot()

# density plot:
# ggplot(mdna, aes(Age, fill = Gender)) +
#   geom_density(alpha = 0.5)
```


```{r}
# Ethnicity and Age
boxplot(mdna$Age ~ mdna$Ethnicity)

quantile(subset(mdna, Ethnicity == "Caucasian")$Age)
quantile(subset(mdna, Ethnicity == "Hispanic")$Age)

# Mann-Whitney U test
wilcox.test(mdna$Age ~ mdna$Ethnicity)
```

```{r}
# ggplot2 alternatives

# boxplot:
# ggplot(mdna, aes(x = Ethnicity, y = Age)) +
#   geom_boxplot()

# density plot:
# ggplot(mdna, aes(Age, fill = Ethnicity)) +
#   geom_density(alpha = 0.5)
```


## DNA methylation values

```{r}
# base summary
summary(mdna[, vec_probe])

# limits of all CpG values
range(apply(mdna[, vec_probe], 2, min), 
      apply(mdna[, vec_probe], 2, max))
```

```{r}
# try: 

# plot(apply(mdna[, vec_probe], 2, min), ylim = c(0, 1), pch = 16, col = 2, type = "l")
# lines(apply(mdna[, vec_probe], 2, max), pch = 16, col = 3)
```

The values are given as methylated percentage.

**Question:** Why should we transform the CpG values?

```{r}
# function for the log-odds transformation
logodds_trans <- function(p) log(p / (1 - p))
```

+ Which model uses this transformation?
    - logistic regression

```{r}
# transform the percentage values into log-odds
probe_logodds <- apply(mdna[, vec_probe], 2, logodds_trans)

# create data frame to work with
mdna_logodds <- data.frame(mdna[, 1:4], probe_logodds)
```

Henceforth we will be using the `mdna_logodds` data.

```{r}
# CpG probes correlation with Age
vec_cor_values <- sapply(seq(vec_probe), function(x) cor(mdna_logodds$Age, mdna_logodds[, vec_probe[x]], method = "spearman"))

# look at the correlation values between each probe and Age
plot(vec_cor_values, ylim = c(-1, 1), ylab = "Spearman's Correlation")
abline(h = 0, lty = 2)

# density plot
plot(density(vec_cor_values), xlab = "Spearman's Correlation", xlim = c(-1, 1))

# save the CpG probes ordered by their absolute correlation value
vec_probe_cor <- vec_probe[order(abs(vec_cor_values), decreasing = TRUE)]
```

**Question:** What does having a high positive/negative correlation might indicate?

# Questions

1. Main question: Can we predict one's biological age?
2. Can we predict the age of an individual based on the methylation level of a given CpG probe?
3. Does gender pose an effect in this prediction?
4. Does ethnicity pose an effect in this prediction?


# **Exercise 1:** Simple linear regression

For this exercise, we will study the relation between age and methylation levels from specific probes.

+ Perform a simple linear regression to study the (linear) relationship between Age and a single specific marker (hint: use the function `lm()`)
+ For each case, what is the model's formula (hint: `coef()` or `coefficients()`)
+ Model diagnosis
    - Look at the residuals (hint: `resid()` or `residuals()`) 
    - Assess their normality (hint: `shapiro.test()`)
+ Interpret the results
    - What are the models' slopes?
    - What are the model predictions?
    - What are the predicted ages of someone with, for any inferred marker, methylation levels of 10%, 25%, 50%, 75%, and 90%? (hint: `predict()` with the *newdata* argument)


## First probe

Linear model using the first CpG probe, `r vec_probe[1]`.

```{r}
# CpG: cg17568996


# linear regression: formula as Y ~ x
fit1 <- lm(Age ~ cg17568996, data = mdna_logodds)
fit1

# look at the models' summary
summary(fit1)

plot(mdna_logodds$cg17568996, mdna_logodds$Age)
abline(fit1)
```

**Questions:**

1. How do you interpret the *Estimate* values?
2. Which hypotheses tests are the p-values shown here referring to? What can we infer from the results?
3. What is the difference between the *Multiple R-squared* and *Adjusted R-squared*?

We can look at the model's residuals distribution. Ideally, the distribution should be symmetric around zero (Normal distribution with $\mu = 0$). We can assess this visually, through the use of a density plot, histogram, violin plot, boxplot, or even a quantile-quantile plot (Q-Q plot).

```{r}
# residual analysis
resid(fit1)
fit1$residuals


# density plot
plot(density(fit1$residuals))


# Q-Q plot
qqnorm(fit1$residuals)
qqline(fit1$residuals)
```

Alternatively, we can apply formal hypothesis testing to infer on the residuals' normality.

```{r}
# proposal 1: Shapiro-Wilk normality test
shapiro.test(fit1$residuals)

# proposal 2: Lilliefors (Kolmogorov-Smirnov) normality test
nortest::lillie.test(fit1$residuals)
```

Assuming this linear relation between age and the `r vec_probe[1]` methylation values, what are the model fitted values (i.e., the prediction of the mean response value to the coefficient values used to formulate the model)?

```{r}
# pred_fit1 <- ___


# look at the fitted results
# ___
```

What are the predicted ages of someone with `r vec_probe[1]` methylation levels of 10%, 25%, 50%, 75%, and 90%?

```{r}
# vector with percentages of interest
methylation_interest <- c(0.1, 0.25, 0.5, 0.75, 0.9)
# don't forget to transform the data!


# predictions_fit1 <- ___
```


## Top probe correlated with age

Perform the same analysis with the highest age-correlated probe, `r vec_probe_cor[1]`.

```{r}
# CpG probe: cg22736354

# linear regression
fit2 <- lm(Age ~ cg22736354, data = mdna_logodds)

# summary
summary(fit2)
```

```{r}
# residuals analysis
par(mfrow = c(2,2))
plot(fit2)
par(mfrow = c(1,1))
```

```{r}
methylation_interest <- c(0.1, 0.25, 0.5, 0.75, 0.9)
logodds_trans(methylation_interest)

pred_fit2 <- predict(fit2, 
                     newdata = data.frame(
                       cg22736354 = logodds_trans(methylation_interest)))
pred_fit2
summary(fit2)
plot(mdna_logodds$cg22736354, mdna_logodds$Age)
abline(fit2)
abline(0,1, lty = 2)

# fitted results
# ___
```

```{r}
# predictions_fit2 <- ___
```


## Comparing results

From the results of the two linear models what do you see?

+ Differences in the summary, R-squared, or coefficients?
+ How about the residuals?

```{r}
# try:

par(mfrow = c(1, 2))
plot(mdna_logodds$Age, predict(fit1), xlim = c(0, 100), ylim = c(0, 100), xlab = "Age, years", ylab = "Fitted age, fit1")
abline(lm(predict(fit1) ~ mdna_logodds$Age))
abline(0,1, lty = 2, col = "gray55")
plot(mdna_logodds$Age, predict(fit2), xlim = c(0, 100), ylim = c(0, 100), xlab = "Age, years", ylab = "Fitted age, fit2")
abline(lm(predict(fit2) ~ mdna_logodds$Age))
abline(0,1, lty = 2, col = "gray55")
par(mfrow = c(1, 1))
```


# **Exercise 2:** Multiple linear regression

For this exercise, we will perform the same analysis for the `r vec_probe_cor[1]` probe, only this time the variables for gender and ethnicity will be included in the model.

+ Perform a multiple linear regression
+ Model diagnosis
+ Compare against the previous model


## **2.1** Adding the coefficient for gender

```{r}
# CpG: cg22736354

# define gender as a factor with "female" as reference
mdna_logodds$Gender <- relevel(factor(mdna_logodds$Gender), ref = "F")


# fit the model
fit2_g <- lm(Age ~ cg22736354 + Gender, data = mdna_logodds)


# summary
summary(fit2)
summary(fit2_g)
```

```{r}
# assessing normality of residuals
# ___
```

```{r}
# predictions for age in each gender
# pred_g_female <- ___
# pred_g_male <- ___
```


## **2.2** Adding the coefficient for ethnicity

```{r}
# CpG: cg22736354


# define gender as a factor with the caucasian group as reference level
mdna_logodds$Ethnicity <- relevel(factor(mdna_logodds$Ethnicity), ref = "Caucasian")


# fit the model
fit2_e <- lm(Age ~ cg22736354 + Ethnicity, data = mdna_logodds)


# summary
summary(fit2_e)

plot(mdna_logodds$cg22736354, mdna_logodds$Age, col = mdna_logodds$Ethnicity, pch = 16)
abline(fit2_e)
```

```{r}
# assessing normality of residuals
shapiro.test(fit2_e$residuals)

plot(density(fit2_e$residuals))
```

```{r}
# predictions for age in europeans and mexicans
pred_e_cauc <- predict(fit2_e, newdata = data.frame(
  cg22736354 = mdna_logodds$cg22736354,
  Ethnicity = "Caucasian"))
pred_e_hisp <- predict(fit2_e, newdata = data.frame(
  cg22736354 = mdna_logodds$cg22736354,
  Ethnicity = "Hispanic"))
```

```{r}
# try:

plot(mdna_logodds$Age, predict(fit2_e), xlim = c(0, 100), ylim = c(0, 100), col = mdna_logodds$Ethnicity, pch = 16)
abline(lm(predict(fit2_e) ~ mdna_logodds$Age))
abline(0,1, lty = 2)

plot(mdna_logodds$Age, predict(fit2_e), xlim = c(0, 100), ylim = c(0, 100), col = mdna_logodds$Ethnicity, pch = 16)
abline(0,1, lty = 2)
abline(lm(pred_e_cauc ~ mdna_logodds$Age), lwd = 3, col = 1)
abline(lm(pred_e_hisp ~ mdna_logodds$Age), lwd = 3, col = 2)
```


## **2.3** Interaction Interaction `cg22736354*Ethnicity`

+ Try to add an interaction term between the two covariates used in the model
+ Estimate the values for $\beta_0$ and $\beta_1$ on each population group

```{r}
# adding ethnicity interaction
fit3 <- lm(Age ~ cg22736354*Ethnicity, data = mdna_logodds)


# summary
summary(fit3)
```

```{r}
# predictions for age in the two groups
# ___
```

**Question:** What is the difference between this model and two independent simple linear regression models for each group?


# Extending the analysis

```{r}
# run a simple linear model on every CpG probe available


# define empty data frame
# simple_lm <- data.frame()


# run for loop
# for(i in seq(vec_probe)) {
#   
#   # define formula to use
#   fit_formula <- as.formula(paste0("Age ~ ", vec_probe[i]))
#   # fit linear model
#   fit <- lm(fit_formula, data = mdna_logodds)
#   # model summary
#   fit_summary <- summary(fit)
#   
#   # create a simple data frame with information about the model
#   out <- data.frame(coef = vec_probe[i], 
#                     estimate = fit_summary$coefficients[2,1], 
#                     estimate_p = fit_summary$coefficients[2,4], 
#                     r_rquared = fit_summary$r.squared,
#                     r_rquared_adj = fit_summary$adj.r.squared,
#                     aic = AIC(fit),
#                     bic = BIC(fit)
#                     )
#   
#   # join rows of each model in output
#   simple_lm <- rbind(simple_lm, out)
# }
```
