---
output: 
  html_document:
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
    code_download: true
title: "Can we predict one's biological age?"
subtitle: "Multiple linear regression"
date: "September 16th, 2022"
author:
  - name: Joao Malato
  - name: Nuno Sepulveda
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = TRUE,
                      message = TRUE, 
                      fig.align = "center",
                      fig.asp = 0.618,
                      fig.width = 10,
                      dpi = 200, 
                      out.width = "75%")
```


# Libraries {-}

```{r libraries, message=FALSE, warning=FALSE}
library(here)       # easy directory management, as here("YOUR/FILE/PATH")
library(data.table) # fast data wrangling
library(magrittr)   # use of pipe operator `%>%`
library(ggplot2)    # graphics
library(nortest)    # normality assessment test
library(MASS)       # stepwise regression
```


# Load data

Still using the same subset of whole blood data from 656 human individuals, made available by [Hannun et al. (2013)](https://www.sciencedirect.com/science/article/pii/S1097276512008933?via%3Dihub).


```{r}
# base R
# mdna <- read.csv(here("data/methDNA_data_Hannun_etal_2013.csv"), header = TRUE)
mdna_logodds <- read.csv(here("data/logodds_methDNA_data_Hannun_etal_2013.csv"), header = TRUE)
```

```{r}
# look at first 8 cols from the data
head(mdna_logodds[, 1:8])
```

Variables:

* `GEO.ID`: Unique identifier for each one of the `r nrow(mdna_logodds)` participants
* `Age`: Age of participants, in years
* `Gender`: Gender of each individual, binary values for female (`F`) and male (`M`)
* `Ethnicity`: Ethnic group of each individual, binary identifying European participants (`Caucasian`) and Mexican participants (`Hispanic`)
* `cgXXXXXXXX`: CpG marker, each one indicating the methylation values, in percentage

```{r}
# binary variables to factors with defined reference
mdna_logodds$Gender <- relevel(factor(mdna_logodds$Gender), ref = "F")
mdna_logodds$Ethnicity <- relevel(factor(mdna_logodds$Ethnicity), ref = "Caucasian")
```

```{r}
# get a vector with all column names with their pattern starting with "cg" (i.e., the names of all CpG probes)
vec_probe <- grep("^cg", colnames(mdna_logodds), value = TRUE)

# ordered by correlation with age
vec_cor_values <- sapply(seq(vec_probe), function(x) cor(mdna_logodds$Age, mdna_logodds[, vec_probe[x]], method = "spearman"))
# save the CpG probes ordered by their absolute correlation value
vec_probe_cor <- vec_probe[order(abs(vec_cor_values), decreasing = TRUE)]
```


# Questions

1. Main question: Can we predict one's biological age?
2. What if we want to add more variables to our model?
3. Which markers should be added? More than one? All of them? Which ones?
4. Is gender significant in out model?
5. Is the participants' ethnic group significant in out model?
6. What makes a good/better model?


# **Exercise 1:** Which coefficients to add?

+ Fit a model with gender, ethnicity, and the first five CpG probes in the data
+ Fit a similar model but this time include the top 5 age-correlated markers
+ Compare the outputs of the two models
    + Which one is more accurate when predicting the biological age of an individual? (hint: `predict()`)
    + What are the values of AIC for each model? (hint: `AIC()` or `extractAIC()`)


```{r}
# first 5 CpGs in the data:
# cg17568996 + cg02245378 + cg14093936 + cg15642326 + cg24991452

# fit the model
# fit5 <- ___

# summary
# ___
```


```{r}
# top 5 correlated CpGs (decreasing order):
# cg22736354 + cg15804973 + cg06493994 + cg08090640 + cg09809672

# fit the models
# fit5_cor <- ___

# summary
# ___
```


```{r}
# try:

# Visualise the fitted values against the age

# par(mfrow = c(1,2))
# plot(mdna_logodds$Age, predict(fit5), xlim = c(0, 100), ylim = c(0, 100), xlab = "Age, years", main = "fit5")
# abline(lm(predict(fit5) ~ mdna_logodds$Age))
# abline(0, 1, col = "gray55", lty = 2)
# plot(mdna_logodds$Age, predict(fit5_cor), xlim = c(0, 100), ylim = c(0, 100), xlab = "Age, years", main = "fit5_cor")
# abline(lm(predict(fit5_cor) ~ mdna_logodds$Age))
# abline(0, 1, col = "gray55", lty = 2)
# par(mfrow = c(1, 1))


# check the correlation between the fitted values and age
# cor(mdna_logodds$Age, predict(fit5))
# cor(mdna_logodds$Age, predict(fit5_cor))
```


# **Exercise 2:** Model with all variables

+ Fit a model with all the variables available in the dataset (with the exception of the individual ID column) (hint: to facilitate you can add a 'dot' to the right side of the linear regression formula, as `lm(Age ~ .)`)
+ Compare the outputs of the model with the previous ones?
    + Is the model better?
    + What are the AIC values? (hint: `extractAIC()`)
    + Can you improve the model?

```{r}
# fit the full model
# fit_full <- ___

# summary and model diagnosis
# ___
```


**Questions:**

  1. Are all the coefficients in the model needed?
  2. Is there a way to automate the "selection" of variables to use in the model?


# **Exercise 3:** Stepwise regression

Let's try to automate the process of variable selection. Since we already have a full model, we will also need an null model (regression with only the intercept).

+ Fit a null model (hint: `lm(Age ~ 1)`)

```{r}
# fit a null model
# fit_null <- ___

# summary
# ___
```

**Question:** What is the formula for the null model? What is the value for $\beta_0$?


+ Use the `step()` function (*{stats}* package), or the `stepAIC()` function (*{MASS}* package)
    + Look through the function's help page to better understand how to use it
+ Perform a stepwise selection
    1. Forward selection (hint1: `direction = "forward"`; hint2: in this process you start with the null model and add covariates)
    2. Backwards elimination (hint: `direction = "backward"`; hint2: here you start with the full model and remove covariates)
    3. Combination of both (hint: `direction = "both"`)
+ Compare the models
    + AIC values
    + Do the models chose the same variables? The same number of variables?
    + Which one is better at predicting biological age?


Note that the stepwise regression is a test-based procedure for variable selection.

```{r}
# forward selection
# fit_step_forward <- stepAIC(___)


# backward elimination
# fit_step_backward <- ___


# using both methods iteratively
# fit_step_both <- ___
```


```{r}
# compare results
# ___
```


# Extending the analysis

```{r}
# does it improve with Gender?


# define empty data frame
# gender_lm <- data.frame()


# run for loop
# for(i in seq(vec_probe)) {
#   
#   fit_formula <- as.formula(paste0("Age ~ Gender + ", vec_probe[i]))
#   fit <- lm(fit_formula, data = mdna_logodds)
#   fit_summary <- summary(fit)
#   
#   out <- data.frame(coef = vec_probe[i], 
#                     genderM = fit_summary$coefficients[2,1], 
#                     genderM_p = fit_summary$coefficients[2,4],
#                     estimate = fit_summary$coefficients[3,1], 
#                     estimate_p = fit_summary$coefficients[3,4], 
#                     r_rquared = fit_summary$r.squared,
#                     r_rquared_adj = fit_summary$adj.r.squared,
#                     aic = AIC(fit),
#                     bic = BIC(fit)
#                     )
#   
#   gender_lm <- rbind(gender_lm, out)
# }
```

```{r}
# does it improve with Ethnicity?


# define empty data frame
# ethnicity_lm <- data.frame()


# run for loop
# for(i in seq(vec_probe)) {
#   
#   fit_formula <- as.formula(paste0("Age ~ Ethnicity + ", vec_probe[i]))
#   fit <- lm(fit_formula, data = mdna_logodds)
#   fit_summary <- summary(fit)
#   
#   out <- data.frame(coef = vec_probe[i], 
#                     ethnicityHisp = fit_summary$coefficients[2,1], 
#                     ethnicityHisp_p = fit_summary$coefficients[2,4],
#                     estimate = fit_summary$coefficients[3,1], 
#                     estimate_p = fit_summary$coefficients[3,4], 
#                     r_rquared = fit_summary$r.squared,
#                     r_rquared_adj = fit_summary$adj.r.squared,
#                     aic = AIC(fit),
#                     bic = BIC(fit)
#                     )
#   
#   ethnicity_lm <- rbind(ethnicity_lm, out)
# }
```


## `ifelse("I'm ahead of schedule", DO, SKIP)`

+ Look for any potential influential points (hint: Cook's distance)
+ Would an interaction term further improve our the models?
+ Problems of correlation? Multicollinearity?
